---
  title: "Pipeline of Differential Expression Analysis of LUSC data"
output: 
  html_document: 
  default
github_document: 
  df_print: paged
html_preview: FALSE
keep_html: TRUE
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })    
---
  
  Full pipeline of DEA with Glimma - Comparing gene expression by sample type of lusc - Lung Squamouss Cell Carcinoma

https://www.cbioportal.org/study/summary?id=lusc_tcga_pub


```{r, error=TRUE, message=FALSE, warning=FALSE, purl=FALSE, results='hide'}
## This chunk automatically generates a text .R version of this script when running within knitr.
#input  = knitr::current_input()  # filename of input document
#output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
#try(knitr::purl(input,output,documentation=2,quiet=T), silent = T)
# Avoid duplicate label error of knitr::purl
options(knitr.duplicate.label = 'allow')
# Code to browse the markdown file with renderized images.
knitr::opts_chunk$set(
  fig.path = "figs/DEA_"
)
```



# Installing and Loading Libraries            


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
packages_bioconductor = c("DESeq2", "Glimma", "limma","biomaRt", "genefilter")
#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed from Bioconductor and loaded
package.check <- lapply(packages_bioconductor, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    BiocManager::install(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})
packages_cran = c("DT", "tidyverse","dplyr", "tibble", "stringr", "data.table", "genefilter", "ggrepel")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed from CRAN and loaded
package.check <- lapply(packages_cran, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})
rm(packages_cran, packages_bioconductor, package.check)
```


```{r}
load("../data/lusc_count.RData")
codes.type <- lusc.clini %>% 
  filter(sample.type %in% c("NT", "TP" )) %>%   
  rownames_to_column("samples") %>%
  dplyr::select(samples)  %>%
  as_vector(.)
```

# Running DESeq2 Differential Expression for sample.type


```{r}
# create the DESeqDataSet object
ddsObj <- DESeqDataSetFromMatrix(countData = as.matrix(lusc.count[,codes.type]),
                                 colData = lusc.clini[codes.type,],
                                 design = as.formula(~ sample.type))
ddsObj <- DESeq(ddsObj)
res.shr <- results(ddsObj)
#res.shr <- DESeq2::lfcShrink(ddsObj, coef=2, res=res.DESeq2, type='apeglm')
```

```{r}
summary(res.shr)
```

```{r}
dea.NT.TP <- as.data.frame(res.shr) %>%
  rownames_to_column("symbol") %>% 
  left_join(lusc.annot, by="symbol") %>% 
  dplyr::rename(logFC=log2FoldChange, FDR=padj)
```


```{r}
load("../data/dea.NT.TP.rda")
df.deseq <- dea.NT.TP  %>% filter(abs(logFC) >= 3, FDR < 0.01)
dim(df.deseq)
```


```{r}
genes.DEA.NT.vs.TP.lst <- unique(df.deseq$symbol)
write(genes.DEA.NT.vs.TP.lst,  file = "genes.DEA.NT.vs.TP.lst")
save(dea.NT.TP, file = "../data/dea.NT.TP.rda", compress = T)
rm(df.deseq, genes.DEA.NT.vs.TP.lst)
```



```{r}
# add a column with the names of only the top 10 genes
library(ggrepel)
cutoff <- sort(dea.NT.TP$pvalue)[10]
shrink.deseq.cut <- dea.NT.TP %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = log2(baseMean), y=logFC)) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=0.5) +
  labs(x="log2(baseMean)", y="log Fold Change") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```

```{r}
ggplot(dea.NT.TP) + 
  geom_histogram(aes(x = log2(baseMean)), bins = 10)
```


```{r}
# add a column with the names of only the top 10 genes
cutoff <- sort(dea.NT.TP$pvalue)[10]
shrink.deseq.cut <- dea.NT.TP %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = logFC, y= -log10(FDR))) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=2) +
  labs(x="log Fold Change", y="-log10(FDR)") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```


 ```{r} 
 res.df <- as.data.frame(res.shr) 
 res.df$log10MeanNormCount <- log10(res.df$baseMean + 1) 
 idx <-(rowSums(counts(ddsObj)) > 0) 
   res.df <- res.df[idx,] 
   res.df$padj[is.na(res.df$padj)] <- 1 
  
   status <- as.numeric(res.df$padj < 0.05) 
  
   glMDPlot(res.df[idx,], 
                            xval="baseMean", 
                            yval="log2FoldChange", 
                            counts=counts(ddsObj)[idx,], 
                            display.columns=c("GeneID"), 
                            anno=data.frame(GeneID=rownames(ddsObj)[idx]), 
                            groups = lusc.clini[codes.type, "sample.type"], 
                            side.xlab = "Group", 
                            side.ylab = "Expression (log2)",          
                            samples = lusc.clini[codes.type, "codes"], 
                            status=status, 
                            folder = "MDPlot_lusc.NT.vs.TP", 
                            html = "index", 
                            launch=FALSE) 
  
   rm(res.shr, res.df, idx, status) 
   ``` 
  
  
  ```{r}
# first remove the filtered genes (FDR=NA) and create a -log10(FDR) column
filtTab.deseq <- dea.NT.TP %>%
  filter(FDR < 0.1) %>%  
  filter(abs(logFC) > 1) %>% 
  filter(!is.na(FDR)) %>%
  mutate(`-log10(FDR)` = -log10(FDR))
filtTab.deseq <- filtTab.deseq  %>%
  mutate(`-log10(FDR)`=pmin(`-log10(FDR)`))
filtTab.deseq <- filtTab.deseq[!duplicated(filtTab.deseq$symbol), ]
rownames(filtTab.deseq) <- filtTab.deseq$symbol  
```


```{r}
de <- as.integer(abs(filtTab.deseq$logFC) >= 3 & filtTab.deseq$FDR <= 0.01)
glXYPlot(
  x = filtTab.deseq$logFC,
  y = -log10(filtTab.deseq$pvalue),
  xlab = "logFC",
  ylab = "-log10pvalue (FDR)",
  main = "NT.vs.TP",
  counts = counts(ddsObj)[filtTab.deseq$symbol,],
  groups = lusc.clini[codes.type, "sample.type"],
  side.xlab = "Group",
  side.ylab = "Expression (log2)",
  samples = lusc.clini[codes.type, "codes"],
  status = de,
  side.main = "symbol",
  display.columns = c("symbol", "logFC", "FDR", "ensgene", "description"),
  anno = filtTab.deseq,
  folder = "XYPlot_lusc.NT.vs.TP",
  html = "index",
  launch = F)
rm(de)
```


# Running DESeq2 Differential Expression for metastasis


```{r}
# load("../data/lusc_count.RData")
# Selecting only samples with M0 or M1 of Primary Tumors cases:
codes.metas <- lusc.clini %>% 
  filter(sample.type == "TP") %>%  
  filter(metastasis %in% c("M0", "M1")) %>%    
  rownames_to_column("samples") %>%
  dplyr::select(samples)  %>%
  as_vector(.)
```


```{r}
# create the DESeqDataSet object
ddsObj <- DESeqDataSetFromMatrix(countData = as.matrix(lusc.count[,codes.metas]),
                                 colData = lusc.clini[codes.metas,],
                                 design = ~ metastasis)
ddsObj <- DESeq(ddsObj)
res.shr  <- results(ddsObj)
#res.shr <- DESeq2::lfcShrink(ddsObj, coef=2, res=res.shr, type='apeglm')
```

```{r}
summary(res.shr)
```


```{r}
dea.M0.M1 <- as.data.frame(res.shr) %>%
  rownames_to_column("symbol") %>% 
  left_join(lusc.annot, by="symbol") %>% 
  dplyr::rename(logFC=log2FoldChange, FDR=padj)
```



```{r}
#load("../data/dea.M0.M1.rda")
df.deseq <- dea.M0.M1  %>% filter(abs(logFC) >= 2, FDR < 0.01)
df.deseq <- df.deseq[!is.na(df.deseq$logFC),]
dim(df.deseq)
```


```{r}
genes.DEA.M0.vs.M1.lst <- unique(df.deseq$symbol)
write(genes.DEA.M0.vs.M1.lst,  file = "genes.DEA.M0.vs.M1.lst")
save(dea.M0.M1, file = "../data/dea.M0.M1.rda", compress = T)
rm(df.deseq, genes.DEA.M0.vs.M1.lst)
```


```{r}
# add a column with the names of only the top 10 genes
cutoff <- sort(dea.M0.M1$pvalue)[10]
shrink.deseq.cut <- dea.M0.M1 %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = log2(baseMean), y=logFC)) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=0.5) +
  labs(x="mean of normalised counts", y="log Fold Change") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```



```{r}
# add a column with the names of only the top 10 genes
cutoff <- sort(dea.M0.M1$pvalue)[10]
shrink.deseq.cut <- dea.M0.M1 %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = logFC, y= -log10(FDR))) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=2) +
  labs(x="log Fold Change", y="-log10(FDR)") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```


 ```{r} 
   res.df <- as.data.frame(res.shr) 
   res.df$log10MeanNormCount <- log10(res.df$baseMean + 1) 
   idx <-(rowSums(counts(ddsObj)) > 0) 
   res.df <- res.df[idx,] 
   res.df$padj[is.na(res.df$padj)] <- 1 
  
   status <- as.numeric(res.df$padj < .1) 
  
   glMDPlot(res.df[idx,], 
                            xval="baseMean", 
                            yval="log2FoldChange", 
                            counts=counts(ddsObj)[idx,], 
                            display.columns=c("GeneID"), 
                            anno=data.frame(GeneID=rownames(ddsObj)[idx]), 
                            groups = lusc.clini[codes.metas, "metastasis"], 
                            side.xlab = "Group", 
                            side.ylab = "Expression (log2)", 
                            samples = lusc.clini[codes.metas, "codes"], 
                            status=status, 
                            folder = "MDPlot_lusc.M0.vs.M1", 
                            html = "index", 
                            launch=FALSE) 
  
   ``` 
  
  
  
  ```{r}
# first remove the filtered genes (FDR=NA) and create a -log10(FDR) column
filtTab.deseq <- dea.M0.M1 %>%
  filter(FDR < 0.2) %>%  
  filter(abs(logFC) > 0.5) %>% 
  filter(!is.na(FDR)) %>%
  mutate(`-log10(FDR)` = -log10(FDR))
filtTab.deseq <- filtTab.deseq  %>%
  mutate(`-log10(FDR)`=pmin(`-log10(FDR)`))
filtTab.deseq <- filtTab.deseq[!duplicated(filtTab.deseq$symbol), ]
rownames(filtTab.deseq) <- filtTab.deseq$symbol  
```


```{r}
de <- as.integer(abs(filtTab.deseq$logFC) >= 2 & filtTab.deseq$FDR <= 0.05)
#idx <-(rowSums(counts(ddsObj)) > 0)
#normCounts <- log2(counts(ddsObj))
#filtCounts <- normCounts[filtTab.deseq$symbol,]
glXYPlot(
  x = filtTab.deseq$logFC,
  y = -log10(filtTab.deseq$FDR),
  xlab = "logFC",
  ylab = "-log10pvalue(FDR)",
  main = "M0.vs.M1",
  counts = counts(ddsObj)[filtTab.deseq$symbol,] ,
  groups = lusc.clini[codes.metas, "metastasis"],
  side.xlab = "Group",
  side.ylab = "Expression",
  samples = lusc.clini[codes.metas, "codes"],
  status = de,
  side.main = "symbol",
  display.columns = c("symbol", "logFC", "FDR", "ensgene", "description"),
  anno = filtTab.deseq,
  folder = "XYPlot_lusc.M0.vs.M1",
  html = "index",
  launch = F)
```

lusc.clini.CS2<-(distinct(lusc.clini.CS))
rownames(lusc.clini.CS2) <- lusc.clini.CS2$codes


```{r}
load("../data/lusc_count.RData")
codes.type <- lusc.clini.CS2 %>% 
  filter(cancer_status %in% c("TF", "WT" )) %>%   
  rownames_to_column("samples") %>%
  dplyr::select(samples)  %>%
  as_vector(.)
```

# Running DESeq2 Differential Expression for cancer_status


```{r}
# create the DESeqDataSet object
ddsObj <- DESeqDataSetFromMatrix(countData = as.matrix(lusc.count[,codes.type]),
                                 colData = lusc.clini.CS2[codes.type,],
                                 design = as.formula(~ cancer_status))
ddsObj <- DESeq(ddsObj)
res.shr <- results(ddsObj)
#res.shr <- DESeq2::lfcShrink(ddsObj, coef=2, res=res.DESeq2, type='apeglm')
```

```{r}
summary(res.shr)
```

```{r}
dea.TF.WT <- as.data.frame(res.shr) %>%
  rownames_to_column("symbol") %>% 
  left_join(lusc.annot, by="symbol") %>% 
  dplyr::rename(logFC=log2FoldChange, FDR=padj)
```


```{r}
load("../data/dea.TF.WT.rda")
df.deseq <- dea.TF.WT  %>% filter(abs(logFC) >= 2, FDR < 0.01)
dim(df.deseq)
```


```{r}
genes.DEA.TF.vs.WT.lst <- unique(df.deseq$symbol)
write(genes.DEA.TF.vs.WT.lst,  file = "genes.DEA.TF.vs.WT2.lst")
save(dea.TF.WT, file = "../data/dea.TF.WT.rda", compress = T)
rm(df.deseq, genes.DEA.TF.vs.WT.lst)
```



```{r}
# add a column with the names of only the top 10 genes
library(ggrepel)
cutoff <- sort(dea.TF.WT$pvalue)[10]
shrink.deseq.cut <- dea.TF.WT %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = log2(baseMean), y=logFC)) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=0.5) +
  labs(x="log2(baseMean)", y="log Fold Change") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```

```{r}
ggplot(dea.TF.WT) + 
  geom_histogram(aes(x = log2(baseMean)), bins = 10)
```


```{r}
# add a column with the names of only the top 10 genes
cutoff <- sort(dea.TF.WT$pvalue)[10]
shrink.deseq.cut <- dea.TF.WT %>% 
  mutate(TopGeneLabel=ifelse(pvalue<=cutoff, symbol, ""))
ggplot(shrink.deseq.cut, aes(x = logFC, y= -log10(FDR))) + 
  geom_point(aes(colour=FDR < 0.01), pch=20, size=2) +
  labs(x="log Fold Change", y="-log10(FDR)") + 
  geom_label_repel(aes(label=TopGeneLabel), 
                   seed = 123,
                   max.time = 3,
                   max.iter = Inf,
                   size = 3,
                   box.padding = 2, 
                   max.overlaps = Inf)
rm(shrink.deseq.cut, cutoff)
```


```{r} 
res.df <- as.data.frame(res.shr) 
res.df$log10MeanNormCount <- log10(res.df$baseMean + 1) 
idx <-(rowSums(counts(ddsObj)) > 0) 
res.df <- res.df[idx,] 
res.df$padj[is.na(res.df$padj)] <- 1 

status <- as.numeric(res.df$padj < 0.05) 

glMDPlot(res.df[idx,], 
         xval="baseMean", 
         yval="log2FoldChange", 
         counts=counts(ddsObj)[idx,], 
         display.columns=c("GeneID"), 
         anno=data.frame(GeneID=rownames(ddsObj)[idx]), 
         groups = lusc.clini.CS2[codes.type, "cancer_status"], 
         side.xlab = "Group", 
         side.ylab = "Expression (log2)",          
         samples = lusc.clini.CS2[codes.type, "codes"], 
         status=status, 
         folder = "MDPlot_lusc.TF.vs.WT", 
         html = "index", 
         launch=FALSE) 

rm(res.shr, res.df, idx, status) 
``` 


```{r}
# first remove the filtered genes (FDR=NA) and create a -log10(FDR) column
filtTab.deseq <- dea.TF.WT %>%
  filter(FDR < 0.1) %>%  
  filter(abs(logFC) > 1) %>% 
  filter(!is.na(FDR)) %>%
  mutate(`-log10(FDR)` = -log10(FDR))
filtTab.deseq <- filtTab.deseq  %>%
  mutate(`-log10(FDR)`=pmin(`-log10(FDR)`))
filtTab.deseq <- filtTab.deseq[!duplicated(filtTab.deseq$symbol), ]
rownames(filtTab.deseq) <- filtTab.deseq$symbol  
```


```{r}
de <- as.integer(abs(filtTab.deseq$logFC) >= 2 & filtTab.deseq$FDR <= 0.01)
glXYPlot(
  x = filtTab.deseq$logFC,
  y = -log10(filtTab.deseq$pvalue),
  xlab = "logFC",
  ylab = "-log10pvalue (FDR)",
  main = "TF.vs.WT",
  counts = counts(ddsObj)[filtTab.deseq$symbol,],
  groups = lusc.clini[codes.type, "sample.type"],
  side.xlab = "Group",
  side.ylab = "Expression (log2)",
  samples = lusc.clini[codes.type, "codes"],
  status = de,
  side.main = "symbol",
  display.columns = c("symbol", "logFC", "FDR", "ensgene", "description"),
  anno = filtTab.deseq,
  folder = "XYPlot_lusc.TF.vs.WT",
  html = "index",
  launch = F)
rm(de)
```

write_csv(lusc.annot, path = "../data/lusc_annot.csv")
