---
  title: "R Notebook Preprocessing and Download of Count Data"
output: html_notebook
---
  
  # Installing and Loading Libraries            
  
{r message=FALSE, warning=FALSE, paged.print=FALSE}
packages_bioconductor = c("TCGAbiolinks", "SummarizedExperiment", "DESeq2", "Glimma", "limma","biomaRt", "genefilter")
#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed from Bioconductor and loaded
package.check <- lapply(packages_bioconductor, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    BiocManager::install(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
}) 
packages_cran = c("DT", "tidyverse","dplyr", "tibble", "stringr", "data.table", "genefilter", "ggrepel")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed from CRAN and loaded
package.check <- lapply(packages_cran, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})
rm(packages_cran, packages_bioconductor, package.check)


# Downloading TCGA-LUSC clinical data from Xenabrowser


```{r}
# Survival data
# https://xenabrowser.net/datapages/?dataset=TCGA-lusc.surviival.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443

#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LUSC.survival.tsv"
#destfile <- "../data/lusc_survival.tsv.gz"
#download.file(url, destfile)
lusc.survi <- read_tsv(gzfile("../data/lusc_survival.tsv"))
lusc.survi <- lusc.survi %>% 
  mutate(sample = str_replace_all(sample, "-", ".")) %>% 
  column_to_rownames("sample") %>% 
  rename(status = OS, obs.time = OS.time, patient_id = '_PATIENT')

lusc.survi <- as.data.frame(lusc.survi)
# Transform days to months? / 30.43945
# Clinical data
# https://xenabrowser.net/datapages/?dataset=TCGA-lusc.GDC_phenotype.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443
#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LUSC.GDC_phenotype.tsv.gz"
#destfile <- "../data/lusc_clinical.tsv.gz"
#download.file(url, destfile)
lusc.clini <- read_tsv(gzfile("../data/lusc_clinica.tsv.gz"))
lusc.clini <- lusc.clini %>%
  dplyr::select(c("submitter_id.samples","prior_malignancy.diagnoses","age_at_initial_pathologic_diagnosis", "gender.demographic",
                  "sample_type_id.samples", "pathologic_M", "pathologic_N", "pathologic_T")) %>% 
  rename(sample = 'submitter_id.samples', 
         prior.dx = 'prior_malignancy.diagnoses', 
         age = 'age_at_initial_pathologic_diagnosis', 
         gender = 'gender.demographic',
         sample.type = 'sample_type_id.samples',
         metastasis = 'pathologic_M',
         neoplasm = 'pathologic_N',
         ajcc.stage = 'pathologic_T') %>% 
  mutate(sample.type = str_replace_all(sample.type, "01", "TP") ) %>% 
  mutate(sample.type = str_replace_all(sample.type, "11", "NT") ) %>% 
  filter(sample.type %in% c("TP", "NT")) %>%  
  mutate(sample = str_replace_all(sample, "-", ".")) %>% 
  filter(sample %in% row.names(lusc.survi)) %>% 
  column_to_rownames("sample") %>% 
  mutate_if(is.character,as.factor)

lusc.clini <- cbind(lusc.clini, lusc.survi[rownames(lusc.clini),])
lusc.clini$codes <- rownames(lusc.clini)
rm(lusc.survi)
```


# Downloading TCGA-lusc counts from Xenabrowser

```{r}
# https://xenabrowser.net/datapages/?dataset=TCGA-lusc.htseq_counts.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443
#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LUSC.htseq_counts.tsv.gz"
#destfile <- "../data/lusc_counts.tsv.gz"
#download.file(url, destfile)
lusc.count <- read_tsv(gzfile("../data/lusc_counts.tsv.gz"))
lusc.count <- as.data.frame(lusc.count)
colnames(lusc.count) <- gsub("-", "\\.", colnames(lusc.count))
row.names(lusc.count) <- sub("\\..*", "", lusc.count$Ensembl_ID)
lusc.count$Ensembl_ID <- NULL
lusc.count <- 2^(lusc.count)-1
lusc.count <- round(lusc.count, digits = 0)
rm(url, destfile)
```

# Select anotation dataset

```{r}
# If you haven't already installed devtools...
#install.packages("devtools")
# Use devtools to install the package
#devtools::install_github("stephenturner/annotables")
library(annotables)
lusc.annot <- grch38 %>%
  dplyr::filter(grch38$ensgene %in%  row.names(lusc.count)) %>%
  dplyr::select(ensgene, symbol, description)
```


# Filtering Counts and Clinical data

```{r}
lusc.annot <- lusc.annot[!duplicated(lusc.annot$symbol), ]
lusc.annot.miRNA <-  lusc.annot %>% 
  dplyr::filter(grepl("^microRNA", description))
lusc.annot.miRNA <-  lusc.annot %>% 
  dplyr::filter(grepl("^MIR", symbol))
lusc.count.miRNA <- lusc.count[lusc.annot.miRNA$ensgene,]
rownames(lusc.count.miRNA) <- lusc.annot.miRNA$symbol
lusc.count <- lusc.count[lusc.annot$ensgene,]
rownames(lusc.count) <- lusc.annot$symbol
lusc.clini <- lusc.clini[rownames(lusc.clini) %in% colnames(lusc.count), ]
lusc.clini <- lusc.clini[!is.na(lusc.clini$sample.type),]
lusc.count <- lusc.count[,colnames(lusc.count) %in% rownames(lusc.clini)]
```


```{r}
save(lusc.count, lusc.clini, lusc.annot, file="../data/lusc_count.RData", compress=T)
```

# Conditions files of Sample Type for R-Peridot

```{r}
codes.type <- lusc.clini %>% 
  filter(sample.type %in% c("TP", "NT")) %>%   
  rownames_to_column("samples") %>%
  dplyr::select(samples)  %>%
  as_vector(.)
write.table(lusc.count[,codes.type], "../data/lusc_type.tsv", quote = F, sep = "\t")
## Conditions file
cond <- lusc.clini %>% 
  filter(sample.type %in% c("TP", "NT")) %>%   
  rownames_to_column("samples") %>%
  dplyr::select("samples", "sample.type")
row.names(cond) <- NULL
cond$sample.type <- as.factor(ifelse(cond$sample.type == "NT", "condition0","condition1"))
colnames(cond) <- c("samples", "condition")
write.table(cond, "../data/lusc_type.tsv.conditions", quote = F, sep = "\t", row.names = F)
rm(codes.type, cond)
```


# Conditions files of Metastasis for R-Peridot


```{r}
# Selecting only samples with M0 or M1 of Primary Tumors cases:
codes.metas <- lusc.clini %>% 
  filter(sample.type == "TP") %>%  
  filter(metastasis %in% c("M0", "M1")) %>%    
  rownames_to_column("samples") %>%
  dplyr::select(samples)  %>%
  as_vector(.)
write.table(lusc.count[,codes.metas], "../data/lusc_metas.tsv", quote = F, sep = "\t")
## Conditions file
cond <- lusc.clini %>% 
  filter(sample.type == "TP") %>%  
  filter(metastasis %in% c("M0", "M1")) %>% 
  rownames_to_column("samples") %>%
  dplyr::select("samples", "metastasis")
row.names(cond) <- NULL
cond$metastasis <- as.factor(ifelse(cond$metastasis == "M0", "condition0","condition1"))
colnames(cond) <- c("samples", "condition")
write.table(cond, "../data/lusc_metas.tsv.conditions", quote = F, sep = "\t", row.names = F)
rm(codes.metas, cond)
```